---
title: "Australian Open Data — Four Findings"
subtitle: "A short analytical report · Gaurav Warke"
author: "Gaurav Warke"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    page-layout: full
    embed-resources: true
    fig-width: 8
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(readr); library(dplyr); library(ggplot2); library(scales); library(jsonlite)

root     <- normalizePath("..")
data_dir <- file.path(root, "data")

ink <- "#141311"; cream <- "#EFEDE7"; accent <- "#B5502E"
theme_portfolio <- function() {
  theme_minimal(base_size = 13) +
    theme(plot.background = element_rect(fill = cream, colour = NA),
          panel.background = element_rect(fill = cream, colour = NA),
          panel.grid.minor = element_blank(),
          panel.grid.major = element_line(colour = alpha(ink, 0.08)),
          plot.title = element_text(face = "bold", colour = ink),
          axis.text = element_text(colour = alpha(ink, 0.8)))
}
```

This report reads the same tidy artefacts that power the interactive dashboards
(`data/*.csv`, written by `scripts/run_all.py` from published Australian open
data). Every figure is reproducible: re-run the pipeline, re-render this report.

## Headline numbers

```{r kpis}
retail <- fromJSON(file.path(data_dir, "retail_demand.json"))
churn  <- fromJSON(file.path(data_dir, "business_churn.json"))
gst    <- read_csv(file.path(data_dir, "gst_reconciliation_by_state.csv"), show_col_types = FALSE)

tibble(
  Metric = c("Retail turnover (Jun 2025)", "Retail YoY growth",
             "Businesses nationally", "Business exit rate",
             "GST pool distributed"),
  Value = c(paste0("$", comma(retail$headline$turnover_m), "M"),
            paste0("+", retail$headline$yoy_pct, "%"),
            comma(churn$total_businesses),
            paste0(churn$flows$exit_rate_pct, "%"),
            paste0("$", round(sum(gst$gst_bn), 1), "bn"))
) |> knitr::kable()
```

## 1. Where Australia's businesses are

```{r fig-churn}
d <- read_csv(file.path(data_dir, "business_churn_by_state.csv"), show_col_types = FALSE) |>
  mutate(state = reorder(state, businesses))
ggplot(d, aes(state, businesses, fill = share_pct)) +
  geom_col(width = 0.72) +
  geom_text(aes(label = paste0(comma(businesses), "  (", share_pct, "%)")),
            hjust = -0.05, size = 3.2, colour = ink) +
  coord_flip(clip = "off") +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.22))) +
  scale_fill_gradient(low = alpha(accent, 0.45), high = accent, guide = "none") +
  labs(title = "Business base by state", x = NULL, y = "Businesses") +
  theme_portfolio()
```

NSW and VIC together hold roughly 60% of the 2.73M actively trading businesses —
a concentration any national go-to-market or servicing plan has to reckon with.

## 2. Where the government advertising dollar goes

```{r fig-ads}
d <- read_csv(file.path(data_dir, "govt_ad_spend_by_channel.csv"), show_col_types = FALSE) |>
  mutate(channel = reorder(channel, spend_m))
ggplot(d, aes(channel, spend_m, fill = channel == "Digital")) +
  geom_col(width = 0.72) +
  geom_text(aes(label = paste0("$", spend_m, "M")), hjust = -0.1, size = 3.2, colour = ink) +
  coord_flip(clip = "off") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  scale_fill_manual(values = c(`TRUE` = accent, `FALSE` = alpha(ink, 0.55)), guide = "none") +
  labs(title = "Media placement by channel, 2023-24", x = NULL, y = "Spend ($M)") +
  theme_portfolio()
```

Digital has overtaken television as the single largest channel — the shift every
media plan is now built around.

## 3. Retail demand trend

```{r fig-retail}
d <- read_csv(file.path(data_dir, "retail_demand_series.csv"), show_col_types = FALSE) |>
  mutate(date = as.Date(paste0(period, "-01")))
ggplot(d, aes(date, turnover_m)) +
  geom_line(colour = accent, linewidth = 1.1) + geom_point(colour = accent, size = 2.4) +
  scale_x_date(date_labels = "%b %Y") +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0.08, 0.12))) +
  labs(title = "Seasonally adjusted monthly retail turnover", x = NULL, y = "Turnover ($M)") +
  theme_portfolio()
```

Turnover has risen steadily to $37.9bn/month, up 4.9% year on year — demand is
resilient, which matters for inventory and supply-risk planning.

## 4. How the GST pool is shared

```{r fig-gst}
d <- read_csv(file.path(data_dir, "gst_reconciliation_by_state.csv"), show_col_types = FALSE) |>
  mutate(state = reorder(state, gst_bn))
ggplot(d, aes(state, gst_bn, fill = share_pct)) +
  geom_col(width = 0.72) +
  geom_text(aes(label = paste0("$", gst_bn, "bn")), hjust = -0.1, size = 3.2, colour = ink) +
  coord_flip(clip = "off") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  scale_fill_gradient(low = alpha(accent, 0.45), high = accent, guide = "none") +
  labs(title = "GST distribution by state/territory, 2026-27", x = NULL, y = "GST ($bn)") +
  theme_portfolio()
```

The ~$102bn pool is redistributed on a needs basis: VIC and NSW receive the most
in dollar terms, while smaller jurisdictions receive far more per person.
